A Plasma Torus Around a Young Low-Mass Star

### 2025-03-21 paper/ms.tex
update

\begin{figure}[!tp]
  \centering
  \includegraphics[width=0.99\textwidth]{figures/f2.pdf}
  \captionsetup{labelformat=empty}
  \caption{{\bf Figure 2 (Movie):}  
  Emission from circumstellar plasma orbiting TIC 141146667.
  The online movie,
  \href{https://lgbouma.com/movies/TIC141146667_sixpanel.mp4}{available here},
  shows the spectral evolution over five hours.
  {\bf Panel a:} Average TESS light curve from 5 February 2024 to 26
  February 2024 (UT) folded on the \periodhr\ hour period.  Black
  points are phase-averaged; gray are the raw data.
  {\bf Panel b:} Keck/HIRES H$\alpha$ spectra acquired on 17
  February 2024.  The continuum is set to unity, and the darkest color
  is at twice the continuum to accentuate emission outside the line
  core ($|v/v_{\rm eq}|>1$, for $v_{\rm eq}$=130\,\kms).  While
  emission in the line core originates in the star's chromosphere,
  the sinusoidal emission features are most readily described by a
  warped plasma torus.
  {\bf Panel c:} Individual epochs of Panel b, visible in the
  online movie.  The dotted line shows the time-averaged spectrum,
  $f_{\langle t \rangle}$.
  {\bf Panel d:} As in Panel a, but overplotting the
  median-normalized H$\alpha$ light curve at $|v/v_{\rm eq}|<1$.
  {\bf Panel e:} As in Panel b, after subtracting the time-averaged
  spectrum.  The line core shows H$\alpha$ excesses and decrements
  advancing from the blue to red wings.
  The asymmetric color stretch
  is set to match the dynamic range of the data.
  {\bf Panel f:} Individual epochs of Panel e, visible in the online
  movie.}
  \label{fig:spec}
\end{figure}

@@

Our Keck/HIRES observations are the first reported time-series
spectra of a CPV, and they show that corotating circumstellar plasma clumps 
are present in at least one such star.
This
observation rules out a ``starspot-only'' origin scenario for CPVs,
\cite{Koen2021} because such scenarios have no means of explaining
spectroscopic emission beyond the stellar disk.  Similarly, scenarios in
which the circumstellar material is made only of dust are also ruled
out.  While dust may be present, to explain the H$\alpha$ emission the
clumps must include plasma with a significant population of hydrogen
atoms in the $n$=3 excited state.  
More quantitatively, the observed H$\alpha$ luminosity suggests
characteristic densities and masses for the gaseous component of these
clumps of $n_{\rm H} \sim 10^{11}$-$10^{12}$\,cm$^{-3}$ and $M_{\rm
gas} \sim 10^{17}$\,g (see Section~\ref{subsec:gas}).
Dust, if present, is independently constrained to have a total
mass $10^{15}\,{\rm g} < M_{\rm dust} < 10^{17}\,{\rm g}$ (see
Section~\ref{subsec:dust}).
This circumstellar material could plausibly originate from
three sites: the star, an old and undetected disk, or outgassing rocky
bodies.  This latter possibility would render CPVs as extrasolar analogs
of the Jupiter-Io plasma torus (e.g.~Ref.~\cite{Bagenal1981}).

@@

\begin{figure}[!b]
  \centering
  \includegraphics[width=0.99\textwidth]{figures/sf1.pdf}
  \caption{Photometric evolution of TIC 141146667 near the Keck/HIRES observation (green bar). 
  {\bf Panel a}: TESS simple aperture photometry. 
  The main data gaps were caused by scattered light from the Earth
  (BTJD 3356-3358.5) and Moon (BTJD 3359.5-3360.5).
  Raw two minute data are in gray; black time-averages to ten minute
  sampling.
  {\bf Panels b,c}: Folded light curve before and after spectroscopy.
  Raw two minute data are in gray; black phase-averages to 100 points
  per cycle.
  During BTJD 3352-3356, a state switch occurred near BTJD 3353,
  in which the dip at $\phi$$\approx$0.8 disappeared.
  While the large eclipse feature was present both before and after
  the HIRES sequence, its photometric shape evolved during the data
  gap.
  }
  \label{fig:fulllc}
\end{figure}

@@

Circumstellar H$\alpha$ emission might be sourced either from
scattering of stellar H$\alpha$ photons, or from radiative
recombination.  We neglect scattering because Figures~\ref{fig:spec}
and~\ref{fig:hirescuts} show the amplitude of the circumstellar
H$\alpha$ emission varying by a factor of $\approx$5, in a manner
uncorrelated with any variability in the chromospheric line core.  The
volume emissivity under case B recombination (in which Lyman-series
photons are trapped by high optical depth, so that recombinations
cascade through H$\alpha$ and other Balmer lines) can be written as
\begin{equation}
  j_{\rm H\alpha} = n_{\rm e} n_{\rm p} \alpha^{\rm eff}_{\rm H\alpha}(T) h \nu_{\rm H\alpha},
\end{equation}
where $n_{\rm e}$ and $n_{\rm p}$ are the electron and proton densities,
and $\alpha^{\rm eff}_{\rm H\alpha}(T)$ is the effective recombination
coefficient, i.e.\ the number of recombinations per unit volume, per
unit time.
For hydrogen with 
densities $n_{\rm H}$$\sim$10$^{10}$\,cm$^{-3}$ and 
temperatures between 1,000-10,000\,K,
$\alpha^{\rm eff}_{\rm H\alpha}$ is typically on the order of $10^{-12}$ to
$10^{-13}$\,cm$^3$\,s$^{-1}$ \cite{Hummer1987,Draine2011}.
Neglecting the effects of atoms other than hydrogen, we can assume
an ionization fraction $x$, such that $n_{\rm e} = n_{\rm p} = x
n_{\rm H}$, for $n_{\rm H}$ the hydrogen number density.  Let $L_{\rm
H\alpha} = j_{\rm H\alpha} V$, for $V$ the volume of the emitting
hydrogen.
The luminosity of circumstellar hydrogen emission, $L_{\rm H\alpha}$,
is an observable: our SED fitting routine yields
$L_\star$$\approx$0.012\,$L_\odot$, which gives
$L_{\rm H\alpha},\star$$\approx$1.0$\times$$10^{28}$erg\,s$^{-1}$, and
the luminosity of the clump $L_{\rm H\alpha}$ is of order one tenth
that of the star.
Assuming an emitting sphere of radius $r$, we can write
\begin{equation}
  n_H = \frac{1}{x}
  \left( 
    \frac{L_{\rm H\alpha}}{ \alpha^{\rm eff}_{\rm H\alpha}(T) h \nu_{\rm H\alpha} V }
  \right)^{1/2}.
\end{equation}
Nominal parameters of $r = 0.1 R_\star$, $x=1/2$, and
$\alpha^{\rm eff}_{\rm H\alpha}$=$10^{-13}$\,cm$^3$\,s$^{-1}$ suggest
$n_{\rm H}\approx 5\times10^{11}$\,cm$^{-3}$.
We have assumed a uniform density clump, and so this yields a total
gas mass $M_{\rm gas}\approx 8\times10^{16}$\,g.

@@

\subsection{Upper and Low Bounds on Dust}\phantom{+}
\label{subsec:dust}

>>>>

\begin{figure}[!tp]
  \centering
  \includegraphics[width=0.99\textwidth]{figures/f2.pdf}
  \captionsetup{labelformat=empty}
  \caption{{\bf Figure 2 (Movie):}  
  Emission from circumstellar plasma orbiting TIC 141146667.
  The online movie,
  \href{https://lgbouma.com/movies/TIC141146667_sixpanel.mp4}{available here},
  shows the spectral evolution over five hours.
  {\bf a,} Average TESS light curve from 5 February 2024 to 26
  February 2024 (UT) folded on the \periodhr\ hour period.  Black
  points are phase-averaged; gray are the raw data.
  {\bf b,} Keck/HIRES H$\alpha$ spectra acquired on 17
  February 2024.  The continuum is set to unity, and the darkest color
  is at twice the continuum to accentuate emission outside the line
  core ($|v/v_{\rm eq}|>1$, for $v_{\rm eq}$=130\,\kms).  While
  emission in the line core originates in the star's chromosphere,
  the sinusoidal emission features are most readily described by a
  warped plasma torus.
  {\bf c,} Individual epochs of Panel b, visible in the
  online movie.  The dotted line shows the time-averaged spectrum,
  $f_{\langle t \rangle}$.
  {\bf d,} As in Panel a, but overplotting the
  median-normalized H$\alpha$ light curve at $|v/v_{\rm eq}|<1$.
  {\bf e,} As in Panel b, after subtracting the time-averaged
  spectrum.  The line core shows H$\alpha$ excesses and decrements
  advancing from the blue to red wings.
  The asymmetric color stretch
  is set to match the dynamic range of the data.
  {\bf f,} Individual epochs of Panel e, visible in the online
  movie.}
  \label{fig:spec}
\end{figure}

@@

Our Keck/HIRES observations are the first reported time-series
spectra of a CPV, and they show that corotating circumstellar plasma clumps 
are present around at least one such star.
This
observation rules out a ``starspot-only'' origin scenario for CPVs,
\cite{Koen2021} because such scenarios have no means of explaining
spectroscopic emission beyond the stellar disk.  Similarly, scenarios in
which the circumstellar material is made only of dust are also ruled
out.  While dust may be present, to explain the H$\alpha$ emission the
clumps must include plasma with a significant population of hydrogen
atoms in the $n$=3 excited state.  
More quantitatively, the observed H$\alpha$ luminosity suggests
characteristic densities and masses for the gaseous component of these
clumps of $n_{\rm H} \sim 10^{11}$-$10^{12}$\,cm$^{-3}$ and $M_{\rm
gas} \sim 10^{17}$\,g (see Section~\ref{subsec:gas}).
Dust, if present, is independently constrained to have a total
mass $10^{15}\,{\rm g} < M_{\rm dust} < 10^{17}\,{\rm g}$ (see
Section~\ref{subsec:dust}).
This circumstellar material could plausibly originate from
three sites: the star, an old and undetected disk, or outgassing rocky
bodies.  This latter possibility would render CPVs as extrasolar analogs
of the Jupiter-Io plasma torus (e.g.~Ref.~\cite{Bagenal1981}).

@@

\begin{figure}[!b]
  \centering
  \includegraphics[width=0.99\textwidth]{figures/sf1.pdf}
  \caption{Photometric evolution of TIC 141146667 near the Keck/HIRES observation (green bar). 
  {\bf a,} TESS simple aperture photometry. 
  The main data gaps were caused by scattered light from the Earth
  (BTJD 3356-3358.5) and Moon (BTJD 3359.5-3360.5).
  Raw two minute data are in gray; black time-averages to ten minute
  sampling.
  {\bf b-c,} Folded light curve before and after spectroscopy.
  Raw two minute data are in gray; black phase-averages to 100 points
  per cycle.
  During BTJD 3352-3356, a state switch occurred near BTJD 3353,
  in which the dip at $\phi$$\approx$0.8 disappeared.
  While the large eclipse feature was present both before and after
  the HIRES sequence, its photometric shape evolved during the data
  gap.
  }
  \label{fig:fulllc}
\end{figure}

@@

Circumstellar H$\alpha$ emission might be sourced either from
scattering of stellar H$\alpha$ photons, or from radiative
recombination.  We neglect scattering because Figures~\ref{fig:spec}
and~\ref{fig:hirescuts} show the amplitude of the circumstellar
H$\alpha$ emission varying by a factor of $\approx$5, in a manner
uncorrelated with any variability in the chromospheric line core.  The
volume emissivity under case B recombination (in which Lyman-series
photons are trapped by high optical depth, so that recombinations
cascade through H$\alpha$ and other Balmer lines) can be written as
\begin{equation}
  j_{\rm H\alpha} = n_{\rm e} n_{\rm p} \alpha^{\rm eff}_{\rm H\alpha}(T) h \nu_{\rm H\alpha},
\end{equation}
where $n_{\rm e}$ and $n_{\rm p}$ are the electron and proton densities,
and $\alpha^{\rm eff}_{\rm H\alpha}(T)$ is the effective recombination
coefficient, i.e.\ the number of recombinations per unit volume, per
unit time.
For hydrogen with 
%densities $n_{\rm H}$$\sim$10$^{10}$\,cm$^{-3}$ and 
temperatures between 1,000-10,000\,K,
$\alpha^{\rm eff}_{\rm H\alpha}$ is typically on the order of $10^{-12}$ to
$10^{-13}$\,cm$^3$\,s$^{-1}$ \cite{Hummer1987,Draine2011}.
Neglecting the effects of atoms other than hydrogen, we can assume
an ionization fraction $x$, such that $n_{\rm e} = n_{\rm p} = x
n_{\rm H}$, for $n_{\rm H}$ the hydrogen number density.  Let $L_{\rm
H\alpha} = j_{\rm H\alpha} V$, for $V$ the volume of the emitting
hydrogen.
The luminosity of circumstellar hydrogen emission, $L_{\rm H\alpha}$,
is an observable: our SED fitting routine yields
$L_\star$$\approx$0.012\,$L_\odot$, which gives
$L_{\rm H\alpha},\star$$\approx$1.0$\times$$10^{28}$erg\,s$^{-1}$, and
the luminosity of the clump $L_{\rm H\alpha}$ is of order one tenth
that of the star.
Assuming an emitting sphere of radius $r$, we can write
\begin{equation}
  n_H = \frac{1}{x}
  \left( 
    \frac{L_{\rm H\alpha}}{ \alpha^{\rm eff}_{\rm H\alpha}(T) h \nu_{\rm H\alpha} V }
  \right)^{1/2}.
\end{equation}
Nominal parameters of $r = 0.1 R_\star$, $x=1/2$, and
$\alpha^{\rm eff}_{\rm H\alpha}$=$10^{-13}$\,cm$^3$\,s$^{-1}$ suggest
$n_{\rm H}\approx 5\times10^{11}$\,cm$^{-3}$.
We have assumed a uniform density clump, and so this yields a total
gas mass $M_{\rm gas}\approx 8\times10^{16}$\,g.

@@

\subsection{Upper and Lower Bounds on Dust}\phantom{+}
\label{subsec:dust}
